---
title: "Post_VAST_Analyses"
author: "Matt Robertson"
date: "10/07/2020"
output: github_document
---

# Introduction

Fan and I have decided to divide the methods of this paper into four sections: 

1. Run a multi-species VAST for yellowtail flounder and American plaice
2. Test for the existence of density-dependent habitat selection (DDHS) by examining residuals from a single linear model
3. If the initial test indicates DDHS, create knot specific models to account for it
4. Correlate temperature distribution with either a) raw local density (i.e. distribution) data if no evidence of DDHS, or b) the residuals of knot-specific relationships if there is evidence of DDHS.

I will not detail the methods for the VAST because they have not changed. Furthermore, details about new methods will be given throughout.

# Testing for density-dependent habitat selection

Myers & Stokes (1989) as well as Shackell et al. (2005) tested for the existence of density-dependent habitat selection by examining the relationship between local and global biomass/abundance. They considered the null hypothesis to be that variation in local density responses is independent of habitat suitability (i.e. a single positive linear or exponential relationship). This relationship would indicate that as global biomass increases, so does density at all locations.  

Alternatively, a density dependent habitat selection process (i.e. the alternate hypothesis) would involve local population increasing at a faster rate in marginal habitats than in prime/core habitats. This would indicate that marginal habitats only become occupied when density increases, while core areas are more stable to global biomass changes. Therefore, a single global model of local density vs. global biomass would violate assumptions of normality.

To apply this hypothesis test to the output of a VAST I will first examine the relationship between all local densities (50 knots) and global biomass each year. If there is a positive relationship, and there is no evidence of model misfit (e.g. non-normal residuals would indicate locally varying relationships) this will serve as evidence for the null hypothesis (density-dependent habitat selection does not play a large role in population distribution). If there is no relationship or evidence of model misfit, this will serve as evidence for the alternative hypothesis and we will test whether there are location specific relationships between local density and global biomass. 

## Yellowtail flounder

```{r, echo=FALSE, message=FALSE}
load("converted_50knots_2factors.RData")
load("spatial_list.RData")
load("temp_mat.RData")
load("temp_yr.RData")
library(ggplot2)
library(raster)
library(sp)
```

```{r, echo=TRUE}
local_d<-Save$Report$D_gcy
global_d<-Save$Report$Index_cyl
locs<-Spatial_List$loc_x

yt_density<-local_d[,2,1:33]
yt_global_density<-global_d[,,1][2,1:33]
ampl_density<-local_d[,1,1:33]
ampl_global_density<-global_d[,,1][1,1:33]
```

A linear model initially appears to fit the yellowtail flounder data, although the $R^{2}$ value is quite low (~0.07).

```{r, echo=TRUE, fig.height=4, fig.width=4, fig.cap="Relationship between yellowtail flounder log(local density) and log(global biomass). Solid red line is the predicted relationship from a linear model. There are 50 local density estimates for every one global biomass estimate."}
par(mar=c(4,4,0,0))
plot(log(as.numeric(yt_density))~log(rep(yt_global_density, each=50)), pch=19, xlab="log(global biomass)", ylab="log(local density)")
lm1<-lm(log(as.numeric(yt_density))~log(rep(yt_global_density, each=50)))
abline(a=lm1$coefficients[1], b=lm1$coefficients[2], col="red", lwd=2)

#setwd("C:\\Users\\mroberts\\OneDrive - Memorial University of Newfoundland\\Marine Institute\\Dissertation Plans\\Chapter 1\\VAST Manuscript\\Plots")
#png("yt_linmod.png", width = 4, height = 4, units = 'in', res=300)
par(mar=c(4,4,0,0))
plot(log(as.numeric(yt_density))~log(rep(yt_global_density, each=50)), pch=19, xlab="log(global biomass)", ylab="log(local density)")
lm1<-lm(log(as.numeric(yt_density))~log(rep(yt_global_density, each=50)))
abline(a=lm1$coefficients[1], b=lm1$coefficients[2], col="red", lwd=2)
#dev.off()
```


```{r, echo=FALSE}
summary(lm1)
```

When examining the residual histogram we can see a multimodal distribution. This not only means that the residuals aren't normally distributed but also that there appear to be two unique distributions underlying the poor fit. We see further evidence of model misspecification in the qq plots as well.

```{r, echo=FALSE,fig.height=4, fig.width=4, fig.cap="Residuals from the yellowtail flounder linear model."}
par(mar=c(4,4,0,0))
hist(lm1$residuals, xlab="Residuals", main="")

#setwd("C:\\Users\\mroberts\\OneDrive - Memorial University of Newfoundland\\Marine Institute\\Dissertation Plans\\Chapter 1\\VAST Manuscript\\Plots")
#png("yt_linmod_hist.png",width = 4, height = 4, units = 'in', res=300)
par(mar=c(4,4,0,0))
hist(lm1$residuals, xlab="Residuals", main="")
#dev.off()
```

```{r, echo=FALSE,fig.height=6, fig.width=6, fig.cap="Model diagnostic plots from the yellowtail flounder linear model."}
par(mfrow = c(2,2))
par(mar=c(4,4,3,3))
plot(lm1)
```

## American plaice

American plaice initally appears to have a more apparent linear trend than we had observed for yellowtail flounder. This is further indicated by the higher $R^2$ value of ~0.38.

```{r, echo=TRUE, fig.width=4, fig.height=4, fig.cap="Relationship between yellowtail flounder log(local density) and log(global biomass). Solid red line is the predicted relationship from a linear model. There are 50 local density estimates for every one global biomass estimate."}
par(mar=c(4,4,0,0))
plot(log(as.numeric(ampl_density))~log(rep(ampl_global_density, each=50)), pch=19, xlab="log(global biomass)", ylab="log(local density)")
lm1<-lm(log(as.numeric(ampl_density))~log(rep(ampl_global_density, each=50)))
abline(a=lm1$coefficients[1], b=lm1$coefficients[2], col="red", lwd=2)

#setwd("C:\\Users\\mroberts\\OneDrive - Memorial University of Newfoundland\\Marine Institute\\Dissertation Plans\\Chapter 1\\VAST Manuscript\\Plots")
#png("ampl_linmod.png", width = 4, height = 4, units = 'in', res=300)
par(mar=c(4,4,0,0))
par(mar=c(4,4,0,0))
plot(log(as.numeric(ampl_density))~log(rep(ampl_global_density, each=50)), pch=19, xlab="log(global biomass)", ylab="log(local density)")
lm1<-lm(log(as.numeric(ampl_density))~log(rep(ampl_global_density, each=50)))
abline(a=lm1$coefficients[1], b=lm1$coefficients[2], col="red", lwd=2)
#dev.off()
```

```{r, echo=FALSE}
summary(lm1)
```

Unlike yellowtail flounder, the residuals for American plaice appear normally distributed with no indication of model misspecification.

```{r, echo=FALSE,fig.height=4, fig.width=4, fig.cap="Residuals from the American plaice linear model."}
par(mar=c(4,4,0,0))
hist(lm1$residuals, xlab="Residuals", main="")

#setwd("C:\\Users\\mroberts\\OneDrive - Memorial University of Newfoundland\\Marine Institute\\Dissertation Plans\\Chapter 1\\VAST Manuscript\\Plots")
#png("ampl_linmod_hist.png",width = 4, height = 4, units = 'in', res=300)
par(mar=c(4,4,0,0))
hist(lm1$residuals, xlab="Residuals", main="")
#dev.off()
```


```{r, echo=FALSE, fig.height=6, fig.width=6, fig.cap="Model diagnostic plots from the American plaice linear model."}
par(mfrow = c(2,2))
par(mar=c(4,4,3,3))
plot(lm1)
```

## Conclusions

This simple test appears to indicate that yellowtail flounder does not fit to the null hypothesis of proportional increase in local and global biomass while American plaice does fit the null hypothesis. Therefore indicating that American plaice does not appear to have density-dependent habitat selection, while yellowtail flounder may. Myers & Stokes (1989) had the same conclusion for American plaice and Simpson & Walsh (2004) had a similar conclusion (using different methods) for yellowtail flounder.

# Testing for locally varying relationships

To examine the potential for density-dependent habitat selection in yellowtail flounder, I created a non-linear random effects model in TMB using, $$\hat{y}_{k}=a_kx^{b_k}$$ to examine local variability in the density-dependent habitat selection relationship (Myers & Stokes, 1989). Where $\hat{y}_{k}$ is the log of knot density +10, with the +10 ensuring that all local densities are positive, and $x$ representing the log of total biomass +10. Both $a_k$ and $b_k$ are random effects to provide unique estimates for all knots.

We would expect that the estimate of $b_K$ for locations would be shallowest/concave in prime/core habitats and highest/convex in marginal habitats. Weaker responses (i.e. shallow $b_k$) would specifically indicate that a location is stable with regional biomass changes.

```{r, error=TRUE, echo=FALSE}

yt_local_d<-log(yt_density)+10
grid_cell<-seq(from=0, to=49, by=1)
yt_global_d<-log(yt_global_density)+10
nyr=33

load("yt_re_output.RData")

report_obj <- save_list$rep_obj
sd.rep<-save_list$sd_rep_obj

b_mat<-matrix(nrow=50, ncol=4)
b_mat[,1] = exp(sd.rep$value)
b_mat[,2] =  sd.rep$sd
b_mat[,3]  = exp(sd.rep$value - qnorm(0.975)*b_mat[,2])
b_mat[,4]  = exp(sd.rep$value + qnorm(0.975)*b_mat[,2])

```


The model converged well, where the gradient of the approximated marginal log-likelihood for all fixed effects was <10-6 and the Hessian matrix was positive definite at the maximum-likelihood estimates. We can observe the specific fits separately here,

```{r, echo=FALSE, fig.cap="Yellowtail flounder local density based on global biomass across the fifty knots (panels) in VAST. The red lines represent the predicted local relationship from the random effects model and grey polygon represents the root mean square error for those relationships.", fig.width=14, fig.height=14}
new_seq<-seq(from=10, to=20, length=100)

resids<-save_list$rep_obj$resid

rmse_mat<-matrix(nrow=50, ncol=100)
rmse_mat2<-matrix(nrow=50, ncol=100)
for(i in 1:50){
rmse_mat[i,1:100]<-exp(report_obj$log_a[i])*new_seq^exp(report_obj$log_b[i])+sqrt(mean((resids[i,])^2))
rmse_mat2[i,1:100]<-exp(report_obj$log_a[i])*new_seq^exp(report_obj$log_b[i])-sqrt(mean((resids[i,])^2))
}

par(mfrow=c(5,10),mar=c(0,0,1,0),oma=c(4,4,4,4))
for(i in 1:50){
plot(yt_local_d[i,]~yt_global_d, pch=19, main=i, xaxt="n", yaxt="n")
polygon(c(new_seq, rev(new_seq)), c(rmse_mat[i,], rev(rmse_mat2[i,])), col="#ddd8d8", density = 100, angle=90)
lines(new_seq,exp(report_obj$log_a[i])*new_seq^exp(report_obj$log_b[i]), col="red")
points(yt_local_d[i,]~yt_global_d, pch=19)
}

par(mfrow=c(5,10),mar=c(0,0,1,0),oma=c(4,4,4,4))
for(i in 1:50){
plot(yt_local_d[i,]~yt_global_d, pch=19, main=i, xaxt="n", yaxt="n")
polygon(c(new_seq, rev(new_seq)), c(rmse_mat[i,], rev(rmse_mat2[i,])), col="#ddd8d8", density = 100, angle=90)
points(yt_local_d[i,]~yt_global_d, pch=19)
lines(new_seq,exp(report_obj$log_a[i])*new_seq^exp(report_obj$log_b[i]), col="red")

}
```


We can then examine the spatial trends for the slopes. Again, we would expect higher slopes (and/or concave relationships) for marginal habitats and shallow slopes (and/or convex relationships) for optimal habitats. Figure 3.2 shows that the shallowest slopes are close to the southeast shoal with values increasing with increased distance from that area. Therefore indicating that ideal habitat for yellowtail flounder is on the southeast shoal. Again, this matches the findings of Simpson & Walsh (2004).

```{r, fig.cap="The estimate of $b_k$ from the knot specific exponential models comparing yellowtail flounder local density based on global biomass across the fifty knots in the VAST.", echo=FALSE}
yt_slopes_df<-data.frame(long=locs[,1], lat=locs[,2], slopes_yt=exp(report_obj$log_b))
yt_slopes_df$id<-seq(from=1, to=50, by=1)

ggplot(yt_slopes_df, aes(x=long, y=lat, fill=slopes_yt))+
  geom_point(size=3, pch=21, colour="black")+
  scale_fill_gradient2(midpoint=1, low="blue", mid="white",
                       high="red")+
  labs(fill='b') +
  xlab("Eastings (km)")+ ylab("Northings (km)")+
  theme_classic()

ggplot(yt_slopes_df, aes(x=long, y=lat, fill=slopes_yt))+
  geom_point(size=3, pch=21, colour="black")+
  scale_fill_gradient2(midpoint=1, low="blue", mid="white",
                       high="red")+
  labs(fill='b') +
  xlab("Eastings (km)")+ ylab("Northings (km)")+
  theme_classic()


```

We can also examine whether the slopes are truly significantly different than 1 and/or 0.

```{r, fig.width=5, fig.height=5, fig.cap="Model estimates of $b_k$ at each knot. The segments represent +/- 1.96*SD. The solid red line indicates 1, the dashed black line indicates 0.", echo=FALSE}
par(mfrow=c(1,1),mar=c(4,4,0,0))
plot(seq(from=1, to=50, by=1), b_mat[,1], pch=19, ylim=c(-5,12),
     xlab="Knot", ylab="b")
abline(h=1, col="red", lwd=2)
abline(h=0, lty=2, lwd=2)
segments(x0=seq(from=1, to=50, by=1), y0=b_mat[,3], x1=seq(from=1, to=50, by=1), y1=b_mat[,4])


par(mfrow=c(1,1),mar=c(4,4,0,0))
plot(seq(from=1, to=50, by=1), b_mat[,1], pch=19, ylim=c(-5,12),
     xlab="Knot", ylab="b")
abline(h=1, col="red", lwd=2)
abline(h=0, lty=2, lwd=2)
points(seq(from=1, to=50, by=1), b_mat[,1], pch=19)
segments(x0=seq(from=1, to=50, by=1), y0=b_mat[,3], x1=seq(from=1, to=50, by=1), y1=b_mat[,4])


```

## Conclusions

There were locally varying relationships between local density and global biomass, indicating that yellowtail flounder is influenced by density-dependent habitat selection. Furthermore, it appears that optimal habitat for yellowtail flounder are centered on the southeast shoal and the most marginal habitat is in division 3L.

# Distribution shifts

## Removing the effect of density-dependent habitat selection

Yellowtail flounder distribution appears to be influenced by density-dependent habitat selection. To ensure that our results are not affected by this mechanism we can examine the residuals from this relationship. They should represent deviations from the density-dependent relationship (i.e. the density-independent influence on distribution).

Figure 4.1 shows a general trend of large positive residuals in 3L from 1985-1992, followed by large negative residuals in 3L until ~2004, relatively low residuals until 2010, and then large positive residuals in 3L onwards. Residuals outside of 3L are generally lower and do not tend to show strong temporal correlation. Therefore indicating that the largest variations from the predictions from the density-dependent habitat selection model occur in the marginal habitats which may be most influenced by environmental conditions.

```{r, fig.cap="Yellowtail flounder local density residuals from the density dependent habitat selection model", fig.width=14, fig.height=14, echo=FALSE}

yt_df<-data.frame(long=locs[,1], lat=locs[,2], report_obj$resid)
colnames(yt_df)[3:35]<-seq(from=1985, to=2017, by=1)
yt_df$id<-seq(from=1, to=50, by=1)


resid_df<-reshape(yt_df,
                  direction = "long",
                  varying = list(names(yt_df)[3:35]),
                  v.names = "Resid",
                  idvar = "id",
                  timevar = "Year",
                  times = 1985:2017)

ggplot(resid_df, aes(x=long, y=lat, fill=Resid))+
  geom_point(size=3, pch=21, colour="black")+
  scale_fill_gradient2(midpoint=0, low="blue", mid="white",
                       high="red")+
  labs(fill='Residuals') +
  xlab("Eastings (km)")+ ylab("Northings (km)")+
  theme_classic()+
  facet_wrap(~Year, ncol = 6)



ggplot(resid_df, aes(x=long, y=lat, fill=Resid))+
  geom_point(size=3, pch=21, colour="black")+
  scale_fill_gradient2(midpoint=0, low="blue", mid="white",
                       high="red")+
  labs(fill='Residuals') +
  xlab("Eastings (km)")+ ylab("Northings (km)")+
  theme_classic()+
  facet_wrap(~Year, ncol = 6)



```


## Comparing distributions through space and time

Since American plaice were not influenced by density-dependent habitat selection, and we were able to remove the effect of density-dependent habitat selection from yellowtail flounder, we can directly examine the density-independent distribution of both species. 

First, we will examine self-correlation through time for temperature, and then both fish distributions to understand how the distributions have changed in relative terms. 

Temperature has been highly correlated (>0.6) throughout the entire time-series (Figure 4.2). There are some years where the correlation is lower than average, however it appears that despite mean temperatures changing, the distribution of temperature has remained relatively constant.

```{r, fig.cap="Spearman correlation of temperature distribution with temperature distribution across time", echo=TRUE, message=FALSE}
YR<-YR
yrs<-seq(from=1985, to=2017, by=1)

true_temp<-data.frame(yrs=YR, temp=t(temp_mat))
yrs<-as.data.frame(yrs)

merged_temps<-merge(true_temp, yrs, by="yrs", all.y=TRUE)

new_mat2<-matrix(nrow=35, ncol=35)
new_mat_p<-matrix(nrow=35, ncol=35)
rev_seq<-seq(from=35, to=1)
for(k in 1:35){
  for(i in 1:35){
    new_mat2[rev_seq[i],k]<-cor.test(temp_mat[,i],temp_mat[,k], method="spearman")$estimate
    new_mat_p[rev_seq[i],k]<-cor.test(temp_mat[,i],temp_mat[,k], method="spearman")$p.value
    #if(new_mat_p[rev_seq[i],k]>0.05){new_mat2[rev_seq[i],k]<-NA}
  }
}

new_mat3<-new_mat2
for(i in 1:35){
  new_mat3[i,1:(35-i)]<-NA
}


par(mfrow=c(1,1), mar=c(0,0,0,4))

plot(raster(new_mat3), axes=F,col = heat.colors(12), bty="n", box=FALSE)


par(mfrow=c(1,1), mar=c(0,0,0,4))

plot(raster(new_mat3), axes=F,col = heat.colors(12), bty="n", box=FALSE)


```

Yellowtail flounder show three distinct periods, an initial state that remained constant for about 7 years, followed by new state that was negatively correlated with the inital state and lasted for ~20 years, and the most recent state which again appears correlated with the first state but not with the second (Figure 4.3).

```{r, echo=FALSE, fig.cap="Spearman correlation of yellowtail flounder distribution with yellowtail flounder distribution across time"}

new_mat2<-matrix(nrow=33, ncol=33)
new_mat_p<-matrix(nrow=33, ncol=33)
rev_seq<-seq(from=33, to=1)
for(k in 1:33){
  for(i in 1:33){
    new_mat2[rev_seq[i],k]<-cor.test(as.numeric(yt_df[,3:35][,i]),as.numeric(yt_df[,3:35][,k]), method="spearman")$estimate
    new_mat_p[rev_seq[i],k]<-cor.test(as.numeric(yt_df[,3:35][,i]),as.numeric(yt_df[,3:35][,k]), method="spearman")$p.value
    #if(new_mat_p[i,k]>0.05){new_mat2[i,k]<-NA}
  }
}

new_mat3<-new_mat2
for(i in 1:33){
  new_mat3[i,1:(33-i)]<-NA
}

par(mfrow=c(1,1), mar=c(0,0,0,4))

plot(raster(new_mat3), axes=F,col = heat.colors(12), bty="n", box=FALSE)


par(mfrow=c(1,1), mar=c(0,0,0,4))

plot(raster(new_mat3), axes=F,col = heat.colors(12), bty="n", box=FALSE)

```

American plaice appear to have two distributional states, an initial state that was correlated with itself for 5-7 years and a second that was negatively correlated with the first state and has persisted to this day (Figure 4.4).

```{r, echo=FALSE, fig.cap="Spearman correlation of American plaice distribution with American plaice distribution across time"}

new_mat2<-matrix(nrow=33, ncol=33)
new_mat_p<-matrix(nrow=33, ncol=33)
rev_seq<-seq(from=33, to=1)
for(k in 1:33){
  for(i in 1:33){
    new_mat2[rev_seq[i],k]<-cor.test(ampl_density[,i],ampl_density[,k], method="spearman")$estimate
    new_mat_p[rev_seq[i],k]<-cor.test(ampl_density[,i],ampl_density[,k], method="spearman")$p.value
    #if(new_mat_p[rev_seq[i],k]>0.05){new_mat2[rev_seq[i],k]<-NA}
  }
}

new_mat3<-new_mat2
for(i in 1:33){
  new_mat3[i,1:(33-i)]<-NA
}

par(mfrow=c(1,1), mar=c(0,0,0,4))
plot(raster(new_mat3), axes=F,col = heat.colors(12), bty="n", box=FALSE)

par(mfrow=c(1,1), mar=c(0,0,0,4))

plot(raster(new_mat3), axes=F,col = heat.colors(12), bty="n", box=FALSE)


```


## Comparing distribution shifts with temperature distributions

Since the distribution of temperature has remained relatively contant through time it will serve as a spatial reference. Where positive correlations with fish distributions will indicate that most fish are found in warm water, and negative correlations will indicate that most fish are found in cold water. If fish distributions have non-lagged patterns then each temperature condition (denoted by a mean temperature) should have the same correlation with fish distribution. Therefore, we would predict that extreme temperatures may generate a change in correlation, but that soon after this change, the fish distribution should return when temperatures return.

Yellowtail flounder show a southward distributional shift after 1991 that was maintained until temperatures hit their highest recorded value in 2010 when they moved northwards again (Figure 4.5). This pattern indicates that yellowtail flounder showed a persistent distributional shift following a decline in temperature.

```{r, echo=TRUE}
legend.col <- function(col, lev){
  
  opar <- par
  
  n <- length(col)
  
  bx <- par("usr")
  
  box.cx <- c(bx[2] + (bx[2] - bx[1]) / 1000,
              bx[2] + (bx[2] - bx[1]) / 1000 + (bx[2] - bx[1]) / 50)
  box.cy <- c(bx[3], bx[3])
  box.sy <- (bx[4] - bx[3]) / n
  
  xx <- rep(box.cx, each = 2)
  
  par(xpd = TRUE)
  for(i in 1:n){
    
    yy <- c(box.cy[1] + (box.sy * (i - 1)),
            box.cy[1] + (box.sy * (i)),
            box.cy[1] + (box.sy * (i)),
            box.cy[1] + (box.sy * (i - 1)))
    polygon(xx+0.02, yy, col = col[i], border = "black")
    
  }
  par(new = TRUE)
  plot(0, 0, type = "n",
       ylim = c(min(lev,na.rm=TRUE), max(lev,na.rm=TRUE)),
       yaxt = "n", ylab = "",
       xaxt = "n", xlab = "",
       frame.plot = FALSE)
  axis(side = 4, las = 2, tick = FALSE, line = .75)
  par <- opar
}
```

```{r, fig.height=7, fig.width=7, fig.cap="Spearman correlation of yellowtail flounder distribution (x) with temperature distribution (y) across time. Time-series on the bottom panel represents mean temperature across the Grand Bank during the survey. Years for temperature and fish distribution are not matched 1:1 due to earlier data for temperature and missing years for temperature. The x-axis in the time-series matches the correlation plot.", echo=TRUE, message=FALSE}

new_mat2<-matrix(nrow=35, ncol=33)
new_mat_p<-matrix(nrow=35, ncol=33)
rev_seq<-seq(from=35, to=1)
for(k in 1:33){
  for(i in 1:35){
    new_mat2[rev_seq[i],k]<-cor.test(temp_mat[,i],as.numeric(yt_df[,3:35][,k]), method="spearman")$estimate
    new_mat_p[rev_seq[i],k]<-cor.test(temp_mat[,i],as.numeric(yt_df[,3:35][,k]), method="spearman")$p.value
    #if(new_mat_p[rev_seq[i],k]>0.05){new_mat2[rev_seq[i],k]<-NA}
  }
}

new_mat3<-new_mat2
for(i in 1:29){
  new_mat3[i,1:(29-i)]<-NA
}


par(mfrow=c(1,1))
layout.matrix <- matrix(c(1, 1,
                          2, 2), nrow = 2, ncol = 2, byrow = TRUE)

layout(mat = layout.matrix,
       heights = c(2, 1), # Heights of the two rows
       widths = c(2, 2)) # Widths of the two columns

par(mar=c(0,2,0,4))
image(raster(new_mat3), axes=F,col = heat.colors(12))
legend.col(col = heat.colors(12), lev = new_mat3)
par(mar=c(2,2,0,4))
plot(yrs$yrs, rowMeans(merged_temps[,2:51], na.rm = TRUE), bty="n",pch=19)
abline(v=2011, lwd=2, col="grey", lty=2)
abline(v=1991, lwd=2, col="grey", lty=2)
points(yrs$yrs, rowMeans(merged_temps[,2:51], na.rm = TRUE), pch=19)
lines(yrs$yrs, rowMeans(merged_temps[,2:51], na.rm = TRUE), lwd=2)

par(mfrow=c(1,1))
layout.matrix <- matrix(c(1, 1,
                          2, 2), nrow = 2, ncol = 2, byrow = TRUE)

layout(mat = layout.matrix,
       heights = c(2, 1), # Heights of the two rows
       widths = c(2, 2)) # Widths of the two columns
par(mar=c(0,2,0,4))
image(raster(new_mat3), axes=F,col = heat.colors(12))
legend.col(col = heat.colors(12), lev = new_mat3)
par(mar=c(2,2,0,4))
plot(yrs$yrs, rowMeans(merged_temps[,2:51], na.rm = TRUE), bty="n",pch=19)
segments(x0=2011, y0=0, x1=2011, y1=2.8, lwd=2, col="grey", lty=2)
segments(x0=1991, y0=0, x1=1991, y1=2.8, lwd=2, col="grey", lty=2)
points(yrs$yrs, rowMeans(merged_temps[,2:51], na.rm = TRUE), pch=19)
lines(yrs$yrs, rowMeans(merged_temps[,2:51], na.rm = TRUE), lwd=2)

```

American plaice also show a southward distributional shift after 1991, however that shift has never been fully reversed (Figure 4.6). This pattern indicates that American plaice also showed a persistent distributional shift following the decline in temperature and that their threshold value to return to the north is higher than the threshold for yellowtail flounder.

```{r, fig.height=7, fig.width=7, echo=FALSE, fig.cap="Spearman correlation of American plaice distribution (x) with temperature distribution (y) across time. Time-series on the bottom panel represents mean temperature across the Grand Bank during the survey. Years for temperature and fish distribution are not matched 1:1 due to earlier data for temperature and missing years for temperature.The x-axis in the time-series matches the correlation plot.", echo=TRUE, message=FALSE}
new_mat2<-matrix(nrow=35, ncol=33)
new_mat_p<-matrix(nrow=35, ncol=33)
rev_seq<-seq(from=35, to=1)
for(k in 1:33){
  for(i in 1:35){
    new_mat2[rev_seq[i],k]<-cor.test(temp_mat[,i],ampl_density[,k], method="spearman")$estimate
    new_mat_p[rev_seq[i],k]<-cor.test(temp_mat[,i],ampl_density[,k], method="spearman")$p.value
    #if(new_mat_p[rev_seq[i],k]>0.05){new_mat2[rev_seq[i],k]<-NA}
  }
}

new_mat3<-new_mat2
for(i in 1:29){
new_mat3[i,1:(29-i)]<-NA
}

par(mfrow=c(1,1))
layout.matrix <- matrix(c(1, 1,
                          2, 2), nrow = 2, ncol = 2, byrow = TRUE)

layout(mat = layout.matrix,
       heights = c(2, 1), # Heights of the two rows
       widths = c(2, 2)) # Widths of the two columns

par(mar=c(0,2,0,4))
image(raster(new_mat3), axes=F,col = heat.colors(12))
legend.col(col = heat.colors(12), lev = new_mat3)
par(mar=c(2,2,0,4))
plot(yrs$yrs, rowMeans(merged_temps[,2:51], na.rm = TRUE), bty="n",pch=19)
abline(v=2011, lwd=2, col="grey", lty=2)
abline(v=1991, lwd=2, col="grey", lty=2)
points(yrs$yrs, rowMeans(merged_temps[,2:51], na.rm = TRUE), pch=19)
lines(yrs$yrs, rowMeans(merged_temps[,2:51], na.rm = TRUE), lwd=2)

par(mfrow=c(1,1))
layout.matrix <- matrix(c(1, 1,
                          2, 2), nrow = 2, ncol = 2, byrow = TRUE)

layout(mat = layout.matrix,
       heights = c(2, 1), # Heights of the two rows
       widths = c(2, 2)) # Widths of the two columns

par(mar=c(0,2,0,4))
image(raster(new_mat3), axes=F,col = heat.colors(12))
legend.col(col = heat.colors(12), lev = new_mat3)
par(mar=c(2,2,0,4))
plot(yrs$yrs, rowMeans(merged_temps[,2:51], na.rm = TRUE), bty="n",pch=19)
segments(x0=2011, y0=0, x1=2011, y1=2.8, lwd=2, col="grey", lty=2)
segments(x0=1991, y0=0, x1=1991, y1=2.8, lwd=2, col="grey", lty=2)
points(yrs$yrs, rowMeans(merged_temps[,2:51], na.rm = TRUE), pch=19)
lines(yrs$yrs, rowMeans(merged_temps[,2:51], na.rm = TRUE), lwd=2)



```

# Overall conclusions

Yellowtail flounder appears to be influenced by density-dependent habitat selection while American plaice does not. Ideal habitat for yellowtail flounder is on the southeast shoal with marginal habitat in the north (division 3L). Since American plaice are not affected by density-dependent habitat selection, it does not need to be accounted for in later analyses. Meanwhile, we can examine the spatial distribution of yellowtail flounder without the effects of density-dependent habitat selection by examining the residuals from the density-dependent relationship. These analyses show that both American plaice and yellowtail flounder shifted their distributions following the coldest point in the time-series (1991). Neither species shifted back, even when temperatures returned to their previous values. Yellowtail only shifted back to a similar distribution once temperatures reached their maximum in 2011, while American plaice has still never shifted back. 

# References

Myers, R.A. and Stokes, K. (1989). Density-dependent habitat utilization of groundfish and the improvement of research survey. ICES Committee Meeting D15.

Simpson, M.R. and Walsh, S.J. (2004). Changes in the spatial structure of Grand Bank yellowtail flounder: Testing MacCall's basin hypothesis. Journal of Sea Research, 51(3-4), 199-210.

Shackell, N.L., Frank, K.T., and Brickman, D.W. (2005). Range contraction may not always predict core areas: an example from marine fish. Ecological applications, 15(4), 1440-1449.



